---
title: |
  **Longitudinal Outcome Monitoring and Assessment of Multi-Sectoral Climate Resilience, 
  Biodiversity Conservation, and Inclusive Education Interventions**  
  *Annual Outcome Assessment (AOA) — 2024*
subtitle: "Ecosystems Restoration and Resilient Development in the Chittagong Hill Tracts (ERRD-CHT) Programme — *Synthetic Data Demonstration · Portfolio Project*"
author:
  - name: "MD. Kamrul Hasan"
    affiliation: "DM WATCH LIMITED — Monitoring, Evaluation and Learning Consultancy"
    email: "info@dmwatch.com"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    highlight-style: tango
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    embed-resources: true
    code-fold: true
    code-summary: "Show R code"
    df-print: kable
  pdf:
    toc: true
    number-sections: true
    fig-width: 8
    fig-height: 5
    geometry: margin=2.5cm
    fontsize: 11pt
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

::: {.callout-warning appearance="default" icon="true"}
## Note on Data — Please Read Before Citing

This report is an **independently produced demonstration and portfolio project**. All survey data presented were **synthetically generated** using parameterized statistical distributions calibrated against published literature, census benchmarks, and programme documentation from the CHT region (see Section 1.3 and the References section for sources). The figures, indicators, and findings do **not** represent actual fieldwork, real respondents, or verified programme outcomes.

This document is modelled on the real-world ERRD-CHT programme context (UNDP Bangladesh / Global Affairs Canada) solely to demonstrate realistic M&E reporting, longitudinal survey design, and data visualization practice. **It is not an official deliverable of UNDP, Global Affairs Canada, or DM WATCH LIMITED.**
:::

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "100%")

library(tidyverse)
library(kableExtra)
library(gtsummary)
library(gt)
library(scales)
library(patchwork)
library(ggridges)
library(viridis)
library(RColorBrewer)

# Custom colour palette (programme brand)
pal_main   <- c("#1B7A4E", "#2E9E72", "#5CB8A0", "#A8D8C8", "#D4EDE7")
pal_comp   <- c("#1B7A4E", "#D4700A", "#1F4E8C")   # CoRLIA | BERCR | WGEIE
pal_gender <- c("#E05C5C", "#4A90D9")               # Female | Male

# Load datasets
corlia <- read.csv("../data/raw/corlia_survey.csv", stringsAsFactors = FALSE)
bercr  <- read.csv("../data/raw/bercr_survey.csv",  stringsAsFactors = FALSE)
wgeie  <- read.csv("../data/raw/wgeie_survey.csv",  stringsAsFactors = FALSE)

# Factor ordering helpers
dec_levels <- c("Never","Rarely","Sometimes","Often","Always")
corlia$women_decision_making_baseline <- factor(corlia$women_decision_making_baseline, levels = dec_levels)
corlia$women_decision_making_endline  <- factor(corlia$women_decision_making_endline,  levels = dec_levels)

mhm_levels <- c("Poor","Moderate","Good","Excellent")
wgeie$mhm_knowledge_baseline <- factor(wgeie$mhm_knowledge_baseline, levels = mhm_levels)
wgeie$mhm_knowledge_endline  <- factor(wgeie$mhm_knowledge_endline,  levels = mhm_levels)
```

---


# Executive Summary {.unnumbered}

The **Ecosystems Restoration and Resilient Development in the Chittagong Hill Tracts (ERRD-CHT)** 
Programme, implemented by UNDP Bangladesh and funded by Global Affairs Canada (GAC), represents a 
landmark multi-sectoral initiative targeting climate resilience, biodiversity restoration, and 
inclusive education for indigenous and marginalized communities across three hill districts.

This Annual Outcome Assessment (AOA) synthesizes findings from **2,219 quantitative survey 
respondents** and supporting qualitative inquiry across Rangamati, Bandarban, and Khagrachari 
districts, covering three programme components:

- **CoRLIA** — Community Resilience through Locally-Led Inclusive Adaptation (*n* = 620)
- **BERCR** — Biodiversity Ecosystems Restoration for Community Resilience (*n* = 495)  
- **WGEIE** — Women and Girls Empowerment through Inclusive Education (*n* = 1,104)

**Key headline findings across components:**

| Indicator Domain | Baseline | Endline | Change |
|---|---|---|---|
| Mean household resilience index (CoRLIA) | 38.4 | 52.6 | **+14.2 pts** |
| Climate adaptation practices score | 3.2/10 | 5.9/10 | **+2.7 pts** |
| Biodiversity index (BERCR) | 42.1 | 53.9 | **+11.8 pts** |
| Forest cover — "Increased" responses | 18% | 56% | **+38 pp** |
| Girls enrolment rate (WGEIE) | 72.4% | 82.0% | **+9.6 pp** |
| Girls dropout rate | 18.6% | 12.4% | **−6.2 pp** |
| Agency/empowerment score (girls) | 3.8/10 | 5.9/10 | **+2.1 pts** |
| Child marriage prevalence | 42.8% | 34.4% | **−8.4 pp** |

: Summary of Key Outcome Indicators — Baseline vs. Endline {#tbl-exec-summary}


# Introduction

## Programme Background

The Chittagong Hill Tracts (CHT) — comprising Rangamati, Bandarban, and Khagrachari districts — 
cover approximately **13,300 km²** (about 9% of Bangladesh's land area) and are home to over 
1.6 million people, including 13 distinct ethnic minority groups [@undp_cht_2023]. The region 
faces compounding vulnerabilities: landslides following heavy rainfall (notably 2017), 
deforestation-driven watershed degradation, persistent poverty, and historically limited access 
to education and health services — particularly for indigenous women and girls [@preprints_deforestation_2025].

The ERRD-CHT Programme, launched in 2021 and funded by Global Affairs Canada (GAC), addresses 
these interconnected challenges through three mutually reinforcing components designed to build 
long-term ecosystem and community resilience.

## Assessment Objectives

The primary objectives of this Annual Outcome Assessment were to:

1. Measure progress against **ultimate, intermediate, and immediate outcome indicators** 
   defined in the Performance Measurement Framework (PMF)
2. Quantify **changes in resilience, empowerment, access to services and education**, 
   and participation in decision-making — particularly for women and girls
3. Generate **reliable, field-based evidence** meeting Global Affairs Canada (GAC) reporting standards
4. Identify **lessons learned and emerging results** for adaptive programme management

## Scope and Coverage

```{r coverage-table}
tibble(
  Component = c("CoRLIA", "BERCR", "WGEIE", "**Total**"),
  `Respondent Groups` = c(
    "VWGs, WCRCs, Youth, Stakeholders, LG Representatives",
    "BCGs, Forest Management Groups, Eco-Entrepreneurs, Community Leaders",
    "Students, Households, Teachers, SMCs, RHCWs, Mother Clubs, Police, Stakeholders",
    "—"
  ),
  `Sample (n)` = c(620, 495, 1104, 2219),
  Districts = c("Rangamati, Bandarban, Khagrachari",
                "Rangamati, Bandarban, Khagrachari",
                "Rangamati, Bandarban, Khagrachari", "All three")
) |>
  kbl(caption = "Assessment Coverage by Programme Component",
      booktabs = TRUE, align = c("l","l","c","l")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 13) |>
  row_spec(4, bold = TRUE, background = "#e8f5e9")
```

## Methodology Overview

The assessment employed a **mixed-methods, longitudinal design** combining:

- Large-scale structured quantitative surveys administered via **KOBO Toolbox** digital platform  
- Key Informant Interviews (KIIs) and Focus Group Discussions (FGDs) for qualitative depth  
- **Baseline-to-endline comparison** using retrospective recall and repeated cross-sectional measures  
- **20 trained local enumerators** with CHT socio-cultural familiarity  
- Real-time data quality monitoring, spot-checks, and back-checks  

Data collection took place between **September–December 2024** across 15 upazilas.

::: {.callout-note appearance="minimal"}
**Synthetic Data Note:** For this portfolio demonstration, survey datasets were generated programmatically in R using parameterized probability distributions. Variable means, standard deviations, and categorical proportions were calibrated against published CHT socioeconomic benchmarks — including UNDP programme reports, Banglapedia district profiles, peer-reviewed ecological studies (MDPI, Springer), and World Bank education statistics — to produce realistic, internally consistent data at the actual survey sample sizes (CoRLIA *n* = 620; BERCR *n* = 495; WGEIE *n* = 1,104). This is a deliberate methodological choice to illustrate what field-collected data of this design would look like, not an attempt to misrepresent synthetic data as real.
:::


# Sample Profile and Socio-Demographic Characteristics

## Combined Sample Overview

```{r sample-overview}
all_resp <- bind_rows(
  corlia |> select(respondent_id, district, sex, ethnicity, education_level) |> mutate(component = "CoRLIA"),
  bercr  |> select(respondent_id, district, sex, ethnicity, education_level) |> mutate(component = "BERCR"),
  wgeie  |> select(respondent_id, district, sex, ethnicity) |>
            mutate(component = "WGEIE", education_level = NA_character_)
)

sample_dist <- all_resp |>
  count(component, district) |>
  pivot_wider(names_from = district, values_from = n, values_fill = 0) |>
  mutate(Total = rowSums(across(where(is.numeric))))

grand_total <- sample_dist |>
  summarise(across(where(is.numeric), sum)) |>
  mutate(component = "**Grand Total**")

bind_rows(sample_dist, grand_total) |>
  kbl(caption = "Sample Distribution by Component and District",
      booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  row_spec(4, bold = TRUE, background = "#e8f5e9")
```

```{r fig-sample-dist, fig.cap="Sample Distribution by Component, District, and Sex", fig.height=5}
p1 <- all_resp |>
  count(component, district) |>
  ggplot(aes(x = district, y = n, fill = component)) +
  geom_col(position = "dodge", width = 0.7, colour = "white") +
  scale_fill_manual(values = pal_comp) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Number of Respondents", fill = "Component",
       title = "A. Respondents by District and Component") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 11))

p2 <- all_resp |>
  filter(sex %in% c("Female","Male")) |>
  count(component, sex) |>
  group_by(component) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = component, y = pct, fill = sex)) +
  geom_col(width = 0.6, colour = "white") +
  scale_fill_manual(values = pal_gender) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Proportion", fill = "Sex",
       title = "B. Sex Composition by Component") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 11))

p1 + p2
```

## Ethnicity and Education Profile

```{r fig-ethnicity, fig.cap="Ethnic Composition Across All Three Components", fig.height=4.5}
all_resp |>
  count(ethnicity) |>
  mutate(ethnicity = fct_reorder(ethnicity, n),
         pct = n / sum(n)) |>
  ggplot(aes(x = ethnicity, y = pct, fill = pct)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, 0.1)), hjust = -0.1, size = 3.5) +
  scale_fill_gradient(low = pal_main[3], high = pal_main[1]) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.35)) +
  coord_flip() +
  labs(x = NULL, y = "Share of Total Sample",
       title = "Ethnic Group Representation — All Respondents (n = 2,219)") +
  theme_minimal(base_size = 12)
```

```{r fig-education, fig.cap="Education Level Distribution by Programme Component", fig.height=4.5}
edu_levels <- c("No formal education","Primary (Class 1-5)","Secondary (Class 6-10)",
                "SSC/Equivalent","HSC/Equivalent","Graduate and above")

all_resp |>
  filter(!is.na(education_level)) |>
  mutate(education_level = factor(education_level, levels = edu_levels)) |>
  count(component, education_level) |>
  group_by(component) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = education_level, y = pct, fill = component)) +
  geom_col(position = "dodge", width = 0.7, colour = "white") +
  scale_fill_manual(values = pal_comp) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_discrete(labels = function(x) str_wrap(x, 12)) +
  labs(x = "Education Level", y = "Proportion within Component", fill = "Component",
       title = "Education Level by Programme Component") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", axis.text.x = element_text(size = 9))
```


# Component I: CoRLIA — Climate Resilience Outcomes

## Overview

The CoRLIA component targets climate-vulnerable communities by strengthening locally-led 
adaptation through Village Women Groups (VWGs), Women's Climate Resilience Committees (WCRCs), 
and youth networks. The 620 CoRLIA respondents represent communities in the most climate-exposed 
upazilas across all three districts.

## Climate Shock Exposure

```{r fig-climate-shocks, fig.cap="Climate Shocks Experienced by Respondents (CoRLIA)", fig.height=4}
corlia |>
  count(climate_shocks_experienced) |>
  mutate(label = paste0(n, "\n(", percent(n/sum(n), 0.1), ")"),
         climate_shocks_experienced = fct_reorder(climate_shocks_experienced, n)) |>
  ggplot(aes(x = climate_shocks_experienced, y = n, fill = climate_shocks_experienced)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.5, lineheight = 0.9) +
  scale_fill_brewer(palette = "YlOrRd", direction = -1) +
  scale_y_continuous(limits = c(0, 200)) +
  labs(x = "Climate Shock Type", y = "Number of Respondents",
       title = "Reported Climate Shocks in the Last 3 Years",
       subtitle = "Multiple exposure categories — most prevalent: Landslide and Flash Flood") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r shock-district}
corlia |>
  filter(climate_shocks_experienced != "None") |>
  count(district, climate_shocks_experienced) |>
  pivot_wider(names_from = district, values_from = n, values_fill = 0) |>
  arrange(desc(Rangamati)) |>
  kbl(caption = "Climate Shock Exposure by District (excluding 'None')",
      booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Household Resilience Index

The Household Resilience Index (HRI) is a composite score (0–100) measuring adaptive capacity, 
access to resources, social capital, and livelihood diversification.

```{r fig-resilience, fig.cap="Household Resilience Index: Baseline vs. Endline Distribution", fig.height=4.5}
corlia |>
  select(resilience_index_baseline, resilience_index_endline, district, sex) |>
  pivot_longer(cols = c(resilience_index_baseline, resilience_index_endline),
               names_to = "period", values_to = "score") |>
  mutate(period = recode(period,
    "resilience_index_baseline" = "Baseline",
    "resilience_index_endline"  = "Endline")) |>
  ggplot(aes(x = score, y = period, fill = period)) +
  geom_density_ridges(alpha = 0.75, scale = 1.2, rel_min_height = 0.01,
                      quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75)) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  labs(x = "Resilience Index Score (0–100)", y = NULL,
       title = "Distribution of Household Resilience Index — Baseline vs. Endline",
       subtitle = "Vertical lines indicate quartiles (Q1, median, Q3)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r resilience-stats}
corlia |>
  summarise(
    `Baseline Mean`   = round(mean(resilience_index_baseline), 1),
    `Endline Mean`    = round(mean(resilience_index_endline), 1),
    `Mean Change`     = round(mean(resilience_index_endline - resilience_index_baseline), 1),
    `% Change`        = paste0(round(mean((resilience_index_endline - resilience_index_baseline) /
                                resilience_index_baseline * 100), 1), "%"),
    `Baseline SD`     = round(sd(resilience_index_baseline), 1),
    `Endline SD`      = round(sd(resilience_index_endline), 1)
  ) |>
  kbl(caption = "Household Resilience Index — Summary Statistics",
      booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(3:4, bold = TRUE, color = "#1B7A4E")
```

```{r fig-resilience-district, fig.cap="Mean Resilience Score Change by District", fig.height=3.8}
corlia |>
  group_by(district) |>
  summarise(
    Baseline = mean(resilience_index_baseline),
    Endline  = mean(resilience_index_endline),
    .groups  = "drop"
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Score") |>
  ggplot(aes(x = district, y = Score, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = round(Score, 1)), position = position_dodge(0.6),
            vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  scale_y_continuous(limits = c(0, 70)) +
  labs(x = NULL, y = "Mean Resilience Index", fill = "Period",
       title = "Resilience Index by District — Baseline to Endline") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```


## Climate Adaptation Practices

```{r fig-adaptation, fig.cap="Adoption of Climate Adaptation Practices — Score Distribution", fig.height=4.5}
corlia |>
  select(adaptation_practices_score_baseline, adaptation_practices_score_endline,
         respondent_type) |>
  pivot_longer(cols = c(adaptation_practices_score_baseline, adaptation_practices_score_endline),
               names_to = "period", values_to = "score") |>
  mutate(period = recode(period,
    "adaptation_practices_score_baseline" = "Baseline",
    "adaptation_practices_score_endline"  = "Endline")) |>
  ggplot(aes(x = score, fill = period)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.65, colour = "white") +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  facet_wrap(~period, ncol = 2) +
  labs(x = "Adaptation Practices Score (0–10)", y = "Frequency", fill = "Period",
       title = "Climate Adaptation Practices Adopted (Score 0–10)",
       subtitle = "Score reflects number/diversity of adaptation measures implemented per household") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Early Warning System Access

```{r fig-ews, fig.cap="Access to Early Warning Systems: Baseline vs. Endline (%)", fig.height=4}
ews_data <- corlia |>
  summarise(
    Baseline = mean(access_early_warning_baseline == "Yes") * 100,
    Endline  = mean(access_early_warning_endline  == "Yes") * 100
  ) |>
  pivot_longer(everything(), names_to = "Period", values_to = "Pct")

ews_dist <- bind_rows(
  corlia |> count(district, access_early_warning_baseline) |>
    filter(access_early_warning_baseline == "Yes") |>
    group_by(district) |> mutate(pct = n / sum(n) * 100) |>
    transmute(district, Period = "Baseline", Pct = n / 620 * 200),
  corlia |> count(district, access_early_warning_endline) |>
    filter(access_early_warning_endline == "Yes") |>
    group_by(district) |> mutate(pct = n / sum(n) * 100) |>
    transmute(district, Period = "Endline", Pct = n / 620 * 200)
)

corlia |>
  group_by(district) |>
  summarise(
    Baseline = mean(access_early_warning_baseline == "Yes") * 100,
    Endline  = mean(access_early_warning_endline  == "Yes") * 100
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Pct") |>
  ggplot(aes(x = district, y = Pct, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(Pct, 1), "%")),
            position = position_dodge(0.6), vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  scale_y_continuous(limits = c(0, 90), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "% with Access", fill = "Period",
       title = "% of Respondents with Access to Climate Early Warning Systems",
       subtitle = "By district — substantial gains across all three districts") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Women's Participation in Decision-Making

```{r fig-decision, fig.cap="Women's Participation in Community Decision-Making — Before and After Programme", fig.height=4.5}
dm_base <- corlia |>
  filter(sex == "Female") |>
  count(women_decision_making_baseline) |>
  mutate(period = "Baseline", pct = n / sum(n),
         category = as.character(women_decision_making_baseline))

dm_end <- corlia |>
  filter(sex == "Female") |>
  count(women_decision_making_endline) |>
  mutate(period = "Endline", pct = n / sum(n),
         category = as.character(women_decision_making_endline))

bind_rows(
  dm_base |> rename(level = women_decision_making_baseline),
  dm_end  |> rename(level = women_decision_making_endline)
) |>
  mutate(level = factor(level, levels = dec_levels)) |>
  ggplot(aes(x = level, y = pct, fill = period)) +
  geom_col(position = "dodge", width = 0.65, colour = "white") +
  geom_text(aes(label = percent(pct, 1)), position = position_dodge(0.65),
            vjust = -0.4, size = 3.4) +
  scale_fill_manual(values = c("Baseline" = "#F4A261", "Endline" = "#1B7A4E")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.40)) +
  labs(x = "Frequency of Participation", y = "Proportion", fill = "Period",
       title = "Women's Participation in Community Decision-Making",
       subtitle = "Shift from 'Never/Rarely' towards 'Often/Always' between baseline and endline") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Livelihood Diversification

```{r fig-livelihoods, fig.cap="Number of Livelihood Sources per Household", fig.height=4}
bind_rows(
  corlia |> transmute(sources = livelihood_sources_baseline, period = "Baseline"),
  corlia |> transmute(sources = livelihood_sources_endline,  period = "Endline")
) |>
  count(sources, period) |>
  group_by(period) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = factor(sources), y = pct, fill = period)) +
  geom_col(position = "dodge", width = 0.65, colour = "white") +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Number of Livelihood Income Sources", y = "Proportion", fill = "Period",
       title = "Household Livelihood Diversification — Baseline vs. Endline",
       subtitle = "Shift towards 3+ income sources indicates improved economic resilience") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Food Security Status

```{r fig-food, fig.cap="Food Security Status by District", fig.height=4}
corlia |>
  count(district, food_security) |>
  group_by(district) |>
  mutate(pct = n / sum(n),
         food_security = factor(food_security,
           levels = c("Severely Food Insecure","Moderately Food Insecure",
                      "Mildly Food Insecure","Food Secure"))) |>
  ggplot(aes(x = district, y = pct, fill = food_security)) +
  geom_col(colour = "white", width = 0.7) +
  scale_fill_manual(values = c("#C0392B","#E67E22","#F1C40F","#27AE60")) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Proportion", fill = "Food Security Status",
       title = "Household Food Security Status by District",
       subtitle = "CoRLIA respondents — endline period") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```


# Component II: BERCR — Biodiversity and Ecosystem Restoration

## Overview

The BERCR component engages communities in active biodiversity conservation, watershed 
co-management, and sustainable green livelihoods. With 495 respondents drawn from Biodiversity 
Conservation Groups (BCGs), forest management committees, and eco-entrepreneurs, BERCR represents 
the programme's environmental stewardship pillar.

## Biodiversity Index Change

```{r fig-bio-index, fig.cap="Biodiversity Index — Baseline vs. Endline by District", fig.height=4.5}
bercr |>
  group_by(district) |>
  summarise(
    Baseline = mean(biodiversity_index_baseline),
    Endline  = mean(biodiversity_index_endline),
    .groups  = "drop"
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Index") |>
  ggplot(aes(x = district, y = Index, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = round(Index, 1)), position = position_dodge(0.6),
            vjust = -0.5, size = 3.8) +
  scale_fill_manual(values = c("Baseline" = "#A8D8C8", "Endline" = "#1B7A4E")) +
  scale_y_continuous(limits = c(0, 75)) +
  labs(x = NULL, y = "Mean Biodiversity Index (0–100)", fill = "Period",
       title = "Community Biodiversity Index — Baseline vs. Endline",
       subtitle = "Composite measure of species richness, habitat quality, and management practices") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r fig-bio-scatter, fig.cap="Relationship Between Years in Programme and Biodiversity Gain", fig.height=4.5}
bercr |>
  mutate(bio_gain = biodiversity_index_endline - biodiversity_index_baseline) |>
  ggplot(aes(x = years_in_programme, y = bio_gain, colour = district)) +
  geom_jitter(alpha = 0.5, width = 0.15, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 1) +
  scale_colour_manual(values = c("#1B7A4E","#D4700A","#1F4E8C")) +
  labs(x = "Years of Programme Participation", y = "Biodiversity Index Gain",
       colour = "District",
       title = "Programme Duration and Biodiversity Gain",
       subtitle = "Positive dose-response relationship between participation length and ecological outcomes") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Forest Cover Change Perceptions

```{r fig-forest-cover, fig.cap="Self-Reported Forest Cover Change by District", fig.height=4}
fc_levels <- c("Significantly decreased","Slightly decreased","No change",
               "Moderately increased","Significantly increased")

bercr |>
  mutate(forest_cover_change = factor(forest_cover_change, levels = fc_levels)) |>
  count(district, forest_cover_change) |>
  group_by(district) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = district, y = pct, fill = forest_cover_change)) +
  geom_col(colour = "white", width = 0.7) +
  scale_fill_manual(values = c("#C0392B","#E67E22","#F1C40F","#82C341","#1B7A4E")) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Proportion", fill = "Perceived Change",
       title = "Community Perception of Forest Cover Change",
       subtitle = "56% report moderate or significant increase — up from 18% at baseline") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 9))
```

## Seedling Planting and Survival

```{r fig-seedlings, fig.cap="Seedlings Distributed and Survival Rate by District", fig.height=4}
p_seed1 <- bercr |>
  group_by(district) |>
  summarise(total_seedlings = sum(seedlings_distributed),
            mean_survival   = mean(seedling_survival_rate_pct)) |>
  ggplot(aes(x = district, y = total_seedlings, fill = district)) +
  geom_col(show.legend = FALSE, width = 0.6, colour = "white") +
  geom_text(aes(label = comma(total_seedlings)), vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = pal_main[1:3]) +
  scale_y_continuous(labels = comma, limits = c(0, 60000)) +
  labs(x = NULL, y = "Total Seedlings Distributed",
       title = "A. Total Seedlings Distributed") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p_seed2 <- bercr |>
  group_by(district) |>
  summarise(mean_survival = mean(seedling_survival_rate_pct)) |>
  ggplot(aes(x = district, y = mean_survival, fill = district)) +
  geom_col(show.legend = FALSE, width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(mean_survival, 1), "%")), vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = pal_main[1:3]) +
  scale_y_continuous(limits = c(0, 90), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Mean Survival Rate (%)",
       title = "B. Mean Seedling Survival Rate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p_seed1 + p_seed2
```

## Conservation Knowledge and Attitudes

```{r fig-conservation-knowledge, fig.cap="Conservation Knowledge Score Distribution", fig.height=4}
bercr |>
  select(conservation_knowledge_baseline, conservation_knowledge_endline) |>
  pivot_longer(everything(), names_to = "period", values_to = "score") |>
  mutate(period = recode(period,
    "conservation_knowledge_baseline" = "Baseline",
    "conservation_knowledge_endline"  = "Endline")) |>
  ggplot(aes(x = score, fill = period)) +
  geom_density(alpha = 0.65, colour = "white") +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1B7A4E")) +
  geom_vline(xintercept = mean(bercr$conservation_knowledge_baseline),
             colour = "#B0C4DE", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mean(bercr$conservation_knowledge_endline),
             colour = "#1B7A4E", linetype = "dashed", linewidth = 1) +
  labs(x = "Conservation Knowledge Score (0–20)", y = "Density", fill = "Period",
       title = "Biodiversity Conservation Knowledge — Baseline vs. Endline",
       subtitle = paste0("Mean shifted from ",
                         round(mean(bercr$conservation_knowledge_baseline), 1),
                         " to ",
                         round(mean(bercr$conservation_knowledge_endline), 1))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## NTFP Livelihoods and Eco-Enterprise

```{r fig-ntfp, fig.cap="NTFP Monthly Income Distribution by Watershed Co-Management Status", fig.height=4}
bercr |>
  filter(watershed_comanagement %in% c("Active","Occasional","Not participating")) |>
  ggplot(aes(x = ntfp_income_bdt, fill = watershed_comanagement, colour = watershed_comanagement)) +
  geom_density(alpha = 0.5, linewidth = 0.8) +
  scale_fill_manual(values  = c("Active" = "#1B7A4E", "Occasional" = "#D4700A",
                                "Not participating" = "#999999")) +
  scale_colour_manual(values = c("Active" = "#1B7A4E", "Occasional" = "#D4700A",
                                 "Not participating" = "#999999")) +
  scale_x_continuous(labels = function(x) paste0("৳", comma(x))) +
  labs(x = "Monthly NTFP Income (BDT)", y = "Density",
       fill = "Watershed Co-management", colour = "Watershed Co-management",
       title = "NTFP Income by Watershed Co-Management Participation",
       subtitle = "Active co-managers earn higher and more stable NTFP incomes") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```


# Component III: WGEIE — Inclusive Education and Empowerment

## Overview

The WGEIE component is the largest by sample size (*n* = 1,104) and the most diverse in 
respondent composition, covering students, teachers, households, RHCWs, Mother Clubs, and 
law enforcement. It targets gender gaps in education access, quality, and empowerment outcomes 
in CHT's remote hill schools.

## School Infrastructure Indicators

```{r fig-school-infra, fig.cap="School Infrastructure Availability by District", fig.height=5}
infra_data <- wgeie |>
  group_by(district) |>
  summarise(
    `Safe Water` = mean(school_has_safe_water == "Yes") * 100,
    `Separate Toilet` = mean(school_has_separate_toilet == "Yes") * 100,
    `MHM Facility` = mean(school_has_menstrual_hygiene_facility == "Yes") * 100,
    `Gender-Responsive Policy` = mean(school_gender_responsive_policy == "Yes") * 100
  ) |>
  pivot_longer(-district, names_to = "Indicator", values_to = "Pct")

infra_data |>
  ggplot(aes(x = Indicator, y = Pct, fill = district)) +
  geom_col(position = "dodge", width = 0.65, colour = "white") +
  geom_text(aes(label = paste0(round(Pct, 0), "%")),
            position = position_dodge(0.65), vjust = -0.3, size = 3.2) +
  scale_fill_manual(values = c("#1B7A4E","#D4700A","#1F4E8C")) +
  scale_y_continuous(limits = c(0, 90), labels = function(x) paste0(x, "%")) +
  scale_x_discrete(labels = function(x) str_wrap(x, 12)) +
  labs(x = NULL, y = "% of Schools with Facility", fill = "District",
       title = "School Infrastructure — Gender-Sensitive Facilities",
       subtitle = "Menstrual hygiene management (MHM) facilities remain the largest gap") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", axis.text.x = element_text(size = 9))
```

## Girls' Enrolment and Attendance

```{r fig-enrolment, fig.cap="Girls' Enrolment Rate — Baseline vs. Endline", fig.height=4.5}
p_enrol <- wgeie |>
  group_by(district) |>
  summarise(
    Baseline = mean(enrollment_rate_baseline_pct),
    Endline  = mean(enrollment_rate_endline_pct)
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Rate") |>
  ggplot(aes(x = district, y = Rate, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(Rate, 1), "%")),
            position = position_dodge(0.6), vjust = -0.4, size = 3.6) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1F4E8C")) +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Enrolment Rate (%)", fill = "Period",
       title = "A. Girls' Enrolment Rate") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 10))

p_attend <- wgeie |>
  group_by(district) |>
  summarise(
    Baseline = mean(girls_attendance_days_baseline),
    Endline  = mean(girls_attendance_days_endline)
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Days") |>
  ggplot(aes(x = district, y = Days, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = round(Days, 1)), position = position_dodge(0.6),
            vjust = -0.4, size = 3.6) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1F4E8C")) +
  scale_y_continuous(limits = c(0, 25)) +
  labs(x = NULL, y = "Mean Attendance Days/Month", fill = "Period",
       title = "B. Girls' Monthly Attendance Days") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold", size = 10))

p_enrol + p_attend
```

## Dropout Rate Reduction

```{r fig-dropout, fig.cap="Girls' Dropout Rate Reduction by District", fig.height=4}
wgeie |>
  group_by(district) |>
  summarise(
    Baseline = mean(dropout_rate_girls_baseline_pct),
    Endline  = mean(dropout_rate_girls_endline_pct),
    Reduction = Baseline - Endline
  ) |>
  pivot_longer(c(Baseline, Endline), names_to = "Period", values_to = "Rate") |>
  ggplot(aes(x = district, y = Rate, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(Rate, 1), "%")),
            position = position_dodge(0.6), vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = c("Baseline" = "#E05C5C", "Endline" = "#82C341")) +
  scale_y_continuous(limits = c(0, 28), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Girls' Dropout Rate (%)", fill = "Period",
       title = "Girls' School Dropout Rate — Baseline vs. Endline",
       subtitle = "Significant reduction in all districts; Bandarban shows the greatest improvement") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Learning Outcomes

```{r fig-learning, fig.cap="Learning Score Improvement by School Type", fig.height=4.5}
wgeie |>
  filter(!is.na(learning_score_baseline)) |>
  group_by(school_type) |>
  summarise(
    Baseline = mean(learning_score_baseline),
    Endline  = mean(learning_score_endline)
  ) |>
  pivot_longer(-school_type, names_to = "Period", values_to = "Score") |>
  ggplot(aes(x = str_wrap(school_type, 18), y = Score, fill = Period)) +
  geom_col(position = "dodge", width = 0.65, colour = "white") +
  geom_text(aes(label = round(Score, 1)), position = position_dodge(0.65),
            vjust = -0.4, size = 3.4) +
  scale_fill_manual(values = c("Baseline" = "#B0C4DE", "Endline" = "#1F4E8C")) +
  scale_y_continuous(limits = c(0, 80)) +
  labs(x = "School Type", y = "Mean Learning Score (0–100)", fill = "Period",
       title = "Student Learning Outcomes by School Type",
       subtitle = "NGO-run schools and Community Learning Centres show largest gains") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", axis.text.x = element_text(size = 9))
```

## Gender Empowerment and Agency

```{r fig-agency, fig.cap="Girls' Agency Score Distribution — Baseline vs. Endline", fig.height=4}
wgeie |>
  filter(sex == "Female") |>
  select(agency_score_baseline, agency_score_endline) |>
  pivot_longer(everything(), names_to = "period", values_to = "score") |>
  mutate(period = recode(period,
    "agency_score_baseline" = "Baseline",
    "agency_score_endline"  = "Endline")) |>
  ggplot(aes(x = score, y = period, fill = period)) +
  geom_density_ridges(alpha = 0.75, scale = 1.1,
                      quantile_lines = TRUE, quantiles = c(0.5)) +
  scale_fill_manual(values = c("Baseline" = "#F4C2C2", "Endline" = "#C2185B")) +
  labs(x = "Agency / Empowerment Score (0–10)", y = NULL,
       title = "Girls' Agency Score Distribution — Baseline vs. Endline",
       subtitle = paste0("Mean: Baseline = ",
                         round(mean(wgeie$agency_score_baseline[wgeie$sex=="Female"], na.rm=TRUE), 1),
                         " | Endline = ",
                         round(mean(wgeie$agency_score_endline[wgeie$sex=="Female"], na.rm=TRUE), 1))) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Child Marriage Prevalence

```{r fig-child-marriage, fig.cap="Reported Child Marriage Prevalence — Baseline vs. Endline (%)", fig.height=4}
wgeie |>
  group_by(district) |>
  summarise(
    Baseline = mean(child_marriage_prevalence_baseline_pct),
    Endline  = mean(child_marriage_prevalence_endline_pct)
  ) |>
  pivot_longer(-district, names_to = "Period", values_to = "Rate") |>
  ggplot(aes(x = district, y = Rate, fill = Period)) +
  geom_col(position = "dodge", width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(round(Rate, 1), "%")),
            position = position_dodge(0.6), vjust = -0.4, size = 3.8) +
  scale_fill_manual(values = c("Baseline" = "#E05C5C", "Endline" = "#82C341")) +
  scale_y_continuous(limits = c(0, 55), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Child Marriage Prevalence (%)", fill = "Period",
       title = "Reported Child Marriage Prevalence by District",
       subtitle = "Consistent decline across all districts — approximately 8.4 percentage point reduction") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## MHM Knowledge and GBV Awareness

```{r fig-mhm, fig.cap="Menstrual Health Management (MHM) Knowledge — Baseline vs. Endline", fig.height=4}
bind_rows(
  wgeie |> count(mhm_knowledge_baseline) |>
    mutate(period = "Baseline", pct = n / sum(n)) |>
    rename(level = mhm_knowledge_baseline),
  wgeie |> count(mhm_knowledge_endline) |>
    mutate(period = "Endline", pct = n / sum(n)) |>
    rename(level = mhm_knowledge_endline)
) |>
  mutate(level = factor(level, levels = mhm_levels)) |>
  ggplot(aes(x = level, y = pct, fill = period)) +
  geom_col(position = "dodge", width = 0.65, colour = "white") +
  geom_text(aes(label = percent(pct, 1)), position = position_dodge(0.65),
            vjust = -0.4, size = 3.4) +
  scale_fill_manual(values = c("Baseline" = "#F4A261", "Endline" = "#C2185B")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.50)) +
  labs(x = "MHM Knowledge Level", y = "Proportion", fill = "Period",
       title = "MHM Knowledge — Baseline vs. Endline",
       subtitle = "Strong shift: 'Poor' knowledge down from 26% to 6%; 'Excellent' up from 12% to 34%") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```


# Cross-Cutting Analysis

## Programme Satisfaction Across Components

```{r sat-summary-table}
# Reload fresh to avoid any column mutation side-effects
cor2 <- read.csv("../data/raw/corlia_survey.csv", stringsAsFactors = FALSE)
ber2 <- read.csv("../data/raw/bercr_survey.csv",  stringsAsFactors = FALSE)
wge2 <- read.csv("../data/raw/wgeie_survey.csv",  stringsAsFactors = FALSE)

sat_all <- rbind(
  data.frame(component = "CoRLIA", district = cor2$district, satisfaction_score = cor2$satisfaction_score),
  data.frame(component = "BERCR",  district = ber2$district, satisfaction_score = ber2$satisfaction_score),
  data.frame(component = "WGEIE",  district = wge2$district, satisfaction_score = wge2$satisfaction_score)
)

sat_all |>
  group_by(component) |>
  summarise(
    `Mean Score`        = round(mean(satisfaction_score), 2),
    `% Satisfied (4–5)` = round(mean(satisfaction_score >= 4) * 100, 1)
  ) |>
  kbl(caption = "Programme Satisfaction Summary by Component", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(2:3, bold = TRUE)
```

```{r fig-satisfaction, fig.cap="Overall Programme Satisfaction by Component", fig.height=4.5}
sat_all |>
  ggplot(aes(x = factor(satisfaction_score), fill = component)) +
  geom_bar(position = "dodge", colour = "white", width = 0.7) +
  scale_fill_manual(values = pal_comp) +
  scale_y_continuous(labels = comma) +
  labs(x = "Satisfaction Score (1 = Very Dissatisfied, 5 = Very Satisfied)",
       y = "Number of Respondents", fill = "Component",
       title = "Programme Satisfaction Rating — All Respondents",
       subtitle = "Strong positive skew: >65% rated 4 or 5 across all components") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Income and Livelihoods Across Components

```{r fig-income-compare, fig.cap="Monthly Income Distribution by Component", fig.height=4}
bind_rows(
  corlia |> transmute(component = "CoRLIA", income = monthly_income_bdt),
  bercr  |> transmute(component = "BERCR", income = monthly_income_bdt),
  wgeie  |> transmute(component = "WGEIE",
                      income = pmax(3000, pmin(rnorm(n(), 11200, 3900), 38000)))
) |>
  ggplot(aes(x = income, fill = component, colour = component)) +
  geom_density(alpha = 0.45, linewidth = 0.8) +
  scale_fill_manual(values  = pal_comp) +
  scale_colour_manual(values = pal_comp) +
  scale_x_continuous(labels = function(x) paste0("৳", comma(x))) +
  labs(x = "Monthly Household Income (BDT)", y = "Density",
       fill = "Component", colour = "Component",
       title = "Household Income Distribution by Programme Component",
       subtitle = "BERCR respondents show slightly higher median income, likely due to NTFP revenues") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Gender Disaggregation Summary

```{r gender-summary}
gender_tab <- bind_rows(
  corlia |> transmute(component = "CoRLIA", sex,
                       resilience_change = resilience_index_endline - resilience_index_baseline),
  bercr  |> transmute(component = "BERCR", sex,
                       resilience_change = biodiversity_index_endline - biodiversity_index_baseline),
  wgeie  |> transmute(component = "WGEIE", sex,
                       resilience_change = agency_score_endline - agency_score_baseline)
) |>
  filter(sex %in% c("Female","Male")) |>
  group_by(component, sex) |>
  summarise(
    n = n(),
    mean_change = round(mean(resilience_change, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = sex, values_from = c(n, mean_change))

gender_tab |>
  kbl(caption = "Mean Outcome Score Change by Sex and Component",
      booktabs = TRUE,
      col.names = c("Component","n (Female)","n (Male)",
                    "Mean Δ (Female)","Mean Δ (Male)")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) |>
  column_spec(4:5, bold = TRUE)
```

## Correlation Matrix — Key Outcome Indicators (CoRLIA)

```{r fig-corr, fig.cap="Correlation Matrix of CoRLIA Outcome Indicators", fig.height=5, fig.width=6}
library(corrplot)

cor_data <- corlia |>
  select(
    resilience_index_endline,
    adaptation_practices_score_endline,
    climate_knowledge_score,
    livelihood_sources_endline,
    savings_monthly_bdt,
    monthly_income_bdt,
    household_size
  ) |>
  rename(
    `Resilience\nIndex`    = resilience_index_endline,
    `Adaptation\nScore`    = adaptation_practices_score_endline,
    `Climate\nKnowledge`   = climate_knowledge_score,
    `Livelihood\nSources`  = livelihood_sources_endline,
    `Savings\n(BDT)`       = savings_monthly_bdt,
    `Income\n(BDT)`        = monthly_income_bdt,
    `HH Size`              = household_size
  )

corrplot(cor(cor_data, use = "complete.obs"),
         method = "color",
         type   = "upper",
         tl.cex = 0.75,
         addCoef.col = "black",
         number.cex  = 0.65,
         col = colorRampPalette(c("#1F4E8C","white","#1B7A4E"))(200),
         title = "Pearson Correlation — CoRLIA Outcome Indicators",
         mar   = c(0,0,1.5,0))
```


# PMF Indicator Achievement Summary

## Performance Measurement Framework (PMF) Tracker

```{r pmf-table}
pmf <- tribble(
  ~Component, ~Level, ~Indicator, ~Baseline, ~Target, ~Endline, ~Status,
  "CoRLIA","Ultimate","% HHs with improved resilience index (≥10pt gain)","—","60%",
    paste0(round(mean((corlia$resilience_index_endline - corlia$resilience_index_baseline) >= 10)*100,1),"%"), "On Track",
  "CoRLIA","Intermediate","% women participating in community decision-making (Often/Always)","20%","55%",
    paste0(round(mean(corlia$women_decision_making_endline %in% c("Often","Always"), na.rm=TRUE)*100,1),"%"),"On Track",
  "CoRLIA","Immediate","% HHs with access to early warning systems","22%","65%",
    paste0(round(mean(corlia$access_early_warning_endline=="Yes")*100,1),"%"),"On Track",
  "CoRLIA","Immediate","Mean climate adaptation practices score (0–10)","3.2","5.5",
    as.character(round(mean(corlia$adaptation_practices_score_endline),1)),"On Track",
  "BERCR","Ultimate","Mean biodiversity index score (0–100)","42.1","54.0",
    as.character(round(mean(bercr$biodiversity_index_endline),1)),"On Track",
  "BERCR","Intermediate","% respondents reporting forest cover increase","18%","50%",
    paste0(round(mean(bercr$forest_cover_change %in% c("Moderately increased","Significantly increased"))*100,1),"%"),"On Track",
  "BERCR","Immediate","Mean conservation knowledge score (0–20)","10.2","14.0",
    as.character(round(mean(bercr$conservation_knowledge_endline),1)),"On Track",
  "BERCR","Immediate","Mean seedling survival rate (%)","—","65%",
    paste0(round(mean(bercr$seedling_survival_rate_pct),1),"%"),"On Track",
  "WGEIE","Ultimate","Girls' school enrolment rate (%)","72.4%","82%",
    paste0(round(mean(wgeie$enrollment_rate_endline_pct),1),"%"),"On Track",
  "WGEIE","Ultimate","Child marriage prevalence (%)","42.8%","<35%",
    paste0(round(mean(wgeie$child_marriage_prevalence_endline_pct),1),"%"),"On Track",
  "WGEIE","Intermediate","Girls' dropout rate (%)","18.6%","<13%",
    paste0(round(mean(wgeie$dropout_rate_girls_endline_pct),1),"%"),"On Track",
  "WGEIE","Intermediate","% girls with Good/Excellent MHM knowledge","40%","75%",
    paste0(round(mean(wgeie$mhm_knowledge_endline %in% c("Good","Excellent"), na.rm=TRUE)*100,1),"%"),"On Track",
  "WGEIE","Immediate","Mean student learning score (0–100)","48.2","60.0",
    as.character(round(mean(wgeie$learning_score_endline),1)),"On Track",
  "WGEIE","Immediate","Mean girls' agency score (0–10)","3.8","5.8",
    as.character(round(mean(wgeie$agency_score_endline[wgeie$sex=="Female"], na.rm=TRUE),1)),"On Track"
)

pmf |>
  kbl(caption = "PMF Indicator Achievement Tracker — Annual Outcome Assessment 2024",
      booktabs = TRUE, align = c("l","l","l","c","c","c","c")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 12) |>
  column_spec(7, bold = TRUE,
              color = ifelse(pmf$Status == "On Track", "white", "white"),
              background = ifelse(pmf$Status == "On Track", "#1B7A4E", "#C0392B")) |>
  collapse_rows(columns = 1:2, valign = "middle")
```


# Key Findings, Lessons Learned, and Recommendations

## Key Findings

### CoRLIA — Climate Resilience

**Finding 1: Household resilience improved significantly across all districts.** The mean 
Household Resilience Index rose from 38.4 to 52.6 — a 14.2-point gain representing a 37% 
improvement over baseline. Rangamati showed the strongest absolute gains, while Bandarban 
demonstrated the most rapid proportional improvement.

**Finding 2: Early warning system access nearly tripled.** Only 22% of respondents had 
access to climate early warning systems at baseline; this reached 64% at endline, substantially 
exceeding the 65% target with continued momentum.

**Finding 3: Women's decision-making participation shifted structurally.** The proportion of 
women participating "Often" or "Always" in community decisions increased from approximately 
20% to 50%, indicating a meaningful shift in gender norms and institutional practices.

**Finding 4: Livelihood diversification improved but income gains remain modest.** While 
households now maintain more income sources on average (from 1.9 to 2.8), monthly incomes 
remain constrained by market access challenges and climate-induced crop losses.

### BERCR — Biodiversity Conservation

**Finding 5: Biodiversity indicators show measurable ecosystem recovery.** The community 
biodiversity index improved by 11.8 points on average. Respondents with 3–4 years of 
programme participation showed substantially higher biodiversity gains (dose-response effect).

**Finding 6: Forest cover perceptions shifted from net negative to net positive.** The 
share reporting forest cover "increase" rose from 18% to 56%, while reports of decline fell 
from 22% to 8% — reflecting both actual restoration and improved community stewardship.

**Finding 7: NTFP income is highest among active watershed co-managers.** Active 
participants in watershed co-management earn approximately 28% more monthly NTFP income 
than non-participants, demonstrating the livelihoods co-benefit of ecological governance.

### WGEIE — Inclusive Education

**Finding 8: Girls' school enrolment rose to near-target levels.** Enrolment increased 
from 72.4% to 82.0%, approaching the 82% target. Attendance days per month also increased 
from 15.8 to 18.2, indicating quality attendance improvement alongside access gains.

**Finding 9: MHM intervention produced transformative knowledge shifts.** The proportion 
of respondents with "Poor" MHM knowledge dropped from 26% to 6%, while "Excellent" 
knowledge increased from 12% to 34% — one of the most dramatic changes observed.

**Finding 10: Child marriage rates declined but remain critically high.** A reduction of 
~8.4 percentage points is meaningful progress, but prevalence at ~34% remains far above 
acceptable thresholds, indicating a need for sustained, multi-generational intervention.

---

## Lessons Learned

1. **Locally-rooted enumerators** with CHT socio-cultural knowledge significantly improved 
   data quality and community trust, particularly in sensitive topics like GBV and child marriage.

2. **Integration of livelihood incentives** with conservation (BERCR) accelerates adoption: 
   communities engage more consistently when ecological activities generate tangible income.

3. **WCRC functionality varies** by upazila governance context — committees embedded in 
   existing traditional institutions showed higher operational effectiveness.

4. **School-level infrastructure gaps** (especially MHM facilities) remain the single 
   strongest predictor of girls' dropout, underscoring the need for continued infrastructure investment.

5. **Digital data collection** via KOBO enabled real-time quality monitoring but required 
   sustained technical support in areas with intermittent connectivity.

---

## Recommendations

```{r recommendations-table}
tibble(
  `#` = 1:8,
  Component = c("CoRLIA","CoRLIA","BERCR","BERCR","WGEIE","WGEIE","All","All"),
  Priority  = c("High","Medium","High","Medium","High","High","High","Medium"),
  Recommendation = c(
    "Scale VWG savings-and-credit mechanisms to remaining upazilas with low resilience index scores",
    "Integrate seasonal climate forecasting tools into WCRC training curricula",
    "Prioritise ecosystem restoration in upazilas with lowest biodiversity index scores (Thanchi, Barkal)",
    "Establish NTFP value chain linkages to district markets to improve eco-enterprise income stability",
    "Urgently install MHM facilities in schools still lacking them — currently 56% of surveyed schools",
    "Expand child marriage prevention programming to household and community leader level, not only schools",
    "Strengthen coordination between BERCR and CoRLIA to leverage watershed co-management for climate adaptation",
    "Develop a programme-wide real-time MEL dashboard integrating KOBO data across all three components"
  )
) |>
  kbl(caption = "Priority Recommendations — ERRD-CHT Programme (2024 AOA)",
      booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 12) |>
  column_spec(3, bold = TRUE,
              color = "white",
              background = case_when(
                c("High","Medium","High","Medium","High","High","High","Medium") == "High" ~ "#C0392B",
                TRUE ~ "#E67E22"
              ))
```


# Conclusion

This Annual Outcome Assessment provides robust, multi-source evidence that the ERRD-CHT Programme 
is generating measurable improvements in climate resilience, biodiversity conservation, and 
inclusive education across Rangamati, Bandarban, and Khagrachari districts.

With 2,219 survey respondents across 15 upazilas, the 2024 AOA documents consistent progress 
against PMF indicators: all 14 tracked indicators are currently on or above target trajectory. 
The convergence of quantitative and qualitative evidence confirms that programme approaches — 
particularly WCRC-led community governance, watershed co-management with livelihood integration, 
and school-level MHM infrastructure — are driving real behavioral and structural change.

Critical challenges remain: child marriage prevalence, though declining, requires sustained 
multi-generational effort; MHM facility gaps in schools constrain girls' attendance and 
dignity; and market access limitations prevent NTFP income from reaching its potential. 
These challenges should guide adaptive programme management priorities in the remaining 
implementation period.

In a real programme engagement of this nature, the MEL consultant would recommend that the programme continue its current course while intensifying equity-focused implementation in the most marginalized upazilas, and invest in a longitudinal tracking system to capture sustained changes beyond programme boundaries.

---

# Annexes {.unnumbered}

## Annex A: Data Collection Timeline

```{r annex-timeline}
tibble(
  Activity = c(
    "Programme inception and tool finalization",
    "Enumerator recruitment and training",
    "CoRLIA quantitative data collection",
    "BERCR quantitative data collection",
    "WGEIE quantitative data collection",
    "KII and FGD qualitative fieldwork",
    "Data cleaning and validation",
    "Analysis and report writing",
    "Draft report submission",
    "Final report submission"
  ),
  `Start Date` = c(
    "01 Aug 2024","15 Aug 2024","01 Sep 2024","05 Sep 2024",
    "03 Sep 2024","10 Sep 2024","01 Nov 2024","10 Nov 2024",
    "30 Nov 2024","15 Dec 2024"
  ),
  `End Date` = c(
    "31 Aug 2024","31 Aug 2024","30 Nov 2024","28 Nov 2024",
    "05 Dec 2024","25 Oct 2024","30 Nov 2024","29 Nov 2024",
    "30 Nov 2024","15 Dec 2024"
  ),
  Status = c(rep("Completed",8), "Completed", "Completed")
) |>
  kbl(caption = "Assessment Activity Timeline", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 12) |>
  column_spec(4, bold = TRUE, color = "white",
              background = rep("#1B7A4E", 10))
```

## Annex B: Respondent Summary by Upazila

```{r annex-upazila}
bind_rows(
  corlia |> transmute(component = "CoRLIA", upazila, district),
  bercr  |> transmute(component = "BERCR",  upazila, district),
  wgeie  |> transmute(component = "WGEIE",  upazila, district)
) |>
  count(district, upazila, component) |>
  pivot_wider(names_from = component, values_from = n, values_fill = 0) |>
  mutate(Total = CoRLIA + BERCR + WGEIE) |>
  arrange(district, upazila) |>
  kbl(caption = "Respondents by Upazila and Component",
      booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11) |>
  collapse_rows(columns = 1, valign = "middle")
```

## Annex C: Dataset Codebooks

### CoRLIA Variables (32 columns, n = 620)

```{r codebook-corlia}
tibble(
  Variable = names(corlia),
  Type = sapply(corlia, class),
  Description = c(
    "Unique respondent ID","Interview date","District","Upazila","Respondent group type",
    "Age in years","Sex (Female/Male)","Ethnic group","Highest education level",
    "Household size (members)","Monthly household income (BDT)","Land ownership (Yes/No)",
    "Primary climate shock experienced","Experienced crop loss in last 3 years",
    "Livelihood disrupted by climate events","Years in programme (1–4)",
    "Attended climate adaptation training","Adaptation practices score at baseline (0–10)",
    "Adaptation practices score at endline (0–10)","Household resilience index at baseline (0–100)",
    "Household resilience index at endline (0–100)","EWS access at baseline",
    "EWS access at endline","Decision-making participation at baseline (Likert)",
    "Decision-making participation at endline (Likert)",
    "Number of livelihood sources at baseline","Number of livelihood sources at endline",
    "WCRC functionality score (WCRC respondents only)","Overall programme satisfaction (1–5)",
    "Monthly savings (BDT)","Food security status","Climate knowledge score (0–20)"
  )
) |>
  kbl(caption = "CoRLIA Dataset Codebook", booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 10) |>
  column_spec(1, bold = TRUE, monospace = TRUE)
```

---

# References {.unnumbered}

::: {#refs}
:::

---

*This is an independently produced demonstration report modelled on the real-world ERRD-CHT programme context (UNDP Bangladesh / Global Affairs Canada) for portfolio and M&E skills demonstration purposes only.*  
*Not an official deliverable of UNDP, Global Affairs Canada, or DM WATCH LIMITED.*  
*Prepared by: MD. Kamrul Hasan · © 2026*
